### 2022-05-13 MSC-GUDMUNDSSON-2022.tex
Finished thesis. Implemented final feedback from instructor.

%% Date in english and icelandic
%% NOTE: THIS IS THE DATE OF THE SUPERVISOR'S SIGNATURE!!!!!!
%% Predefined: \MLjan, \MLfeb, \MLmar, ... \MLdec
%\whensigned{day}{month}{year} %day is only used on some formats, but you must put something.
\whensigned{10}{\MLapr}{2022}

@@

%% An abstract is required to be in both Icelandic and English for most degrees.
%% It is considered good form to limit the abstract to a single paragraph in each language,
%%   at 300 words.  Refer to your degree's instructions.
%% Note: Icelandic quotation marks cannot be typeset using "` and "'.  You should use \enquote{}
%% this is probably due to interactions with the MultiLingual macros.
%% TODO: turn this into more sensible macros to avoid confusion --foley
\newML{AbstractText}{
The world is moving at increasing speed away from generating electricity using fossil fuels, toward more environmentally friendly options, such as solar power. Solar power production is volatile and highly dependent on weather conditions, primarily cloud cover. Instability and unpredictability are undesirable in power production since production must match demand. In order to ensure power availability when dependent on solar power, highly polluting and expensive fossil power stations, \enquote{peaker plants}, are kept on standby to react to the instability in solar power. The ability to reliably predict solar power production provides more predictability to power grid operators, allowing for greater efficiency in grid operation. 
Valuable data exists in weather forecasts, which can be harnessed to increase predictability in solar power generation. We use novel deep learning methods, attention, and specifically temporal fusion transformers, to forecast the irradiance at a solar power station, which correlates very strongly with power output. We base this on Belgingur ehf.'s high-resolution weather forecast data, the 24-hour forecast irradiance values for the area of the power station.
Our model achieves a very low average mean absolute percentage error of 13.21\% on the test dataset for one-day forecasts. In addition, the model outputs a confidence level for its predictions providing insight to a grid operator on whether to expect unpredictability in the power output. The model is able to generate more accurate predictions when utilizing forecast data from a wider area around the power station than exclusively data from the power station‘s immediate vicinity.

}  
{
Hraðar og hraðar erum við að færast frá því að framleiða raforku með jarðefnaeldsneyti, í átt að umhverfisvænni mátum til raforkuframleiðslu eins og sólarorku. Sólarorka er óútreiknanleg og mjög háð veðurfari, aðallega skýjahulu. Óstöðugleiki og óútreiknanleiki eru óákjósanlegir þegar kemur að sólarorku, þar sem framleiðsla verður að vera jöfn neyslu. Til að tryggja afhendingu raforku, þegar sólarorka er notuð, reiða framleiðendur sig oft á dýr og mengandi orkuver sem ganga fyrir jarðefnaeldsneyti og standa í viðbragðsstöðu til að bregðast við óvæntum breytingum á sólarorkuframboði. Áreiðanleg geta til að sjá fyrir um sólarorkuframleiðslu gerir stjórnun raforkukerfa einfaldari og útreiknanlegri, sem eykur skilvirkni þeirra.
Verðmæt gögn eru til í formi veðurspáa, sem hægt er að nýta til að gera sólarorkuframleiðslu útreiknanlegri. Við notum nýjar aðferðir á sviði djúpnáms, athygli(e. attention), og þá sérstaklega tímasamrunabylta (e. temporal fusion transformers) til að spá fyrir um sólargeislun við sólarorkuver, sem fylgir náið raforkuframleiðslu orkuversins. Við byggjum þessa vinnu á háupplausnarveðurspám Belgings ehf., 24 klst. geislunarspám fyrir staðsetningu okuversins.
Módelið nær meðal meðaltölugildisprósentuskekkju (e. mean absolute percentage error) upp á 13,21\% á prófunargagnasettinu fyrir eins dags spár. Að auki skilar módelið öryggisbili fyrir spárnar, sem veitir raforkuframleiðendum innsæi um það hvort þeir geti búist við óútreiknanlegri sólarorkuframleiðslu. Módelið skilar nákvæmari spám þegar það styðst við veðurspágögn af stærra svæði í nánd orkuversins en þegar það nýtir eingöngu gögn fyrir nærumhverfi orkuversins.
} % Icelandic abstract goes here
\setAbstract{\MLAbstractText}

@@

%% The list of abbreviations is an example of a special list
%% Other lists may be added, such as lists of algorithms, symbols, theorems, etc.
%% IN CS PhD, this is sometimes centered.
\coverchapter{List of Abbreviations}%%RUM: Not mentioned
\begin{tabular}{ll}
MSc &Master of Science\\
PhD &Doctor of Philosophy\\
AI &Artificial Intelligence\\
ANN &Artificial Neural Network\\
NN &Neural Network\\
RNN &Recurrent Neural Network\\
GRU &Gated Recurrent Unit\\
LSTM &Long Short Term Memory\\
NLP &Natural Language Processing\\
TCNN &Temporal Convolutional Neural Network\\
NARXNN &Non-linear Auto Regressive Neural Network\\
TFT &Temporal Fusion Transformer\\
GRN &Gated Residual Network\\
NWP &Numerical Weather Prediction\\
WRF &Weather Research and Forecasting\\
NCAR &National Center for Atmospheric Research\\
NOAA &National Oceanic and Atmospheric Administration\\
AFWA &Air Force Weather Agency\\
NRL &Naval Research Laboratory\\
OU &University of Oklahoma\\
FAA &Federal Aviation Administration\\
MAPE &Mean Absolute Percentage Error\\
MAE &Mean Absolute Error\\
CUDA &Compute Unified Device Architecture\\
AMD &Advanced Micro Devices inc.\\
CPU &Central Processing Unit\\
RAM &Random Access Memory\\

>>>>

%% Date in english and icelandic
%% NOTE: THIS IS THE DATE OF THE SUPERVISOR'S SIGNATURE!!!!!!
%% Predefined: \MLjan, \MLfeb, \MLmar, ... \MLdec
%\whensigned{day}{month}{year} %day is only used on some formats, but you must put something.
\whensigned{10}{\MLmay}{2022}

@@

%% An abstract is required to be in both Icelandic and English for most degrees.
%% It is considered good form to limit the abstract to a single paragraph in each language,
%%   at 300 words.  Refer to your degree's instructions.
%% Note: Icelandic quotation marks cannot be typeset using "` and "'.  You should use \enquote{}
%% this is probably due to interactions with the MultiLingual macros.
%% TODO: turn this into more sensible macros to avoid confusion --foley
\newML{AbstractText}{
The world is moving at increasing speed away from generating electricity using fossil fuels, toward more environmentally friendly options, such as solar power. Solar power production is volatile and highly dependent on weather conditions, primarily cloud cover. Instability and unpredictability are undesirable in power production since production must match demand. In order to ensure power availability when dependent on solar power, highly polluting and expensive fossil power stations, \enquote{peaker plants}, are kept on standby to react to the instability in solar power. The ability to reliably predict solar power production provides more predictability to power grid operators, allowing for greater efficiency in grid operation. 
Valuable data exists in weather forecasts, which can be harnessed to increase predictability in solar power generation. We use novel deep learning methods, attention, and specifically temporal fusion transformers, to forecast the irradiance at a solar power station, which correlates very strongly with power output. We base this on Belgingur ehf.'s high-resolution weather forecast data, the 24-hour forecast irradiance values for the area of the power station.
Our model achieves excellent results for one-day forecasts on the test dataset, with an average mean absolute percentage error of 13.21\%. In addition, the model outputs a confidence level for its predictions providing insight to a grid operator on whether to expect unpredictability in the power output. The model is able to generate more accurate predictions when utilizing forecast data from a wider area around the power station than exclusively data from the power station‘s immediate vicinity.

}  
{
Heimurinn er að færast frá því að framleiða raforku með jarðefnaeldsneyti, í átt að umhverfisvænni leiðum til raforkuframleiðslu eins og sólarorku. Sólarorka er óútreiknanleg og mjög háð veðurfari, aðallega skýjahulu. Óstöðugleiki og óútreiknanleiki eru óæskilegir þegar kemur að sólarorku, þar sem framleiðsla verður að vera jöfn neyslu. Til að tryggja afhendingu raforku þegar sólarorka er notuð, reiða framleiðendur sig oft á dýr og mengandi orkuver sem ganga fyrir jarðefnaeldsneyti og standa í viðbragðsstöðu til að bregðast við óvæntum breytingum á sólarorkuframboði. Sólarorkuframleiðsla yrði skilvirkari og áreiðanlegri ef hægt væri að spá nákvæmar um framleiðslugetu dagsins, t.d. með áreiðanlegri veðurspám og áhrif þeirra á framleiðslugetu.
Verðmæt gögn eru til í formi veðurspáa sem hægt er að nýta til að gera sólarorkuframleiðslu útreiknanlegri. Við notum nýjar aðferðir á sviði djúpnáms, athygli(e. attention), og þá sérstaklega tímasamrunabylta (e. temporal fusion transformers) til að spá fyrir um sólargeislun við sólarorkuver sem fylgir náið raforkuframleiðslu orkuversins. Við byggjum þessa vinnu á háupplausnarveðurspám Belgings ehf., 24 klst. geislunarspám fyrir staðsetningu okuversins.
Módelið nær meðal meðaltölugildisprósentuskekkju (e. mean absolute percentage error) upp á 13,21\% á prófunargagnasettinu fyrir eins dags spár. Að auki skilar módelið öryggisbili fyrir spárnar, sem veitir raforkuframleiðendum innsæi um það hvort þeir geti búist við óútreiknanlegri sólarorkuframleiðslu. Módelið skilar nákvæmari spám þegar það styðst við veðurspágögn af stærra svæði í nánd orkuversins en þegar það nýtir eingöngu gögn fyrir nærumhverfi orkuversins.
} % Icelandic abstract goes here
\setAbstract{\MLAbstractText}

@@

%% The list of abbreviations is an example of a special list
%% Other lists may be added, such as lists of algorithms, symbols, theorems, etc.
%% IN CS PhD, this is sometimes centered.
\coverchapter{List of Abbreviations}%%RUM: Not mentioned
\begin{tabular}{ll}
MSc &Master of Science\\
PhD &Doctor of Philosophy\\
AI &Artificial Intelligence\\
ANN &Artificial Neural Network\\
NN &Neural Network\\
RNN &Recurrent Neural Network\\
CNN &Convolutional Neural Network\\
GRU &Gated Recurrent Unit\\
LSTM &Long Short Term Memory\\
NLP &Natural Language Processing\\
TCNN &Temporal Convolutional Neural Network\\
NARXNN &Non-linear Auto-Regressive Neural Network\\
TFT &Temporal Fusion Transformer\\
GRN &Gated Residual Network\\
NWP &Numerical Weather Prediction\\
WRF &Weather Research and Forecasting\\
NCAR &National Center for Atmospheric Research\\
NOAA &National Oceanic and Atmospheric Administration\\
AFWA &Air Force Weather Agency\\
NRL &Naval Research Laboratory\\
OU &University of Oklahoma\\
FAA &Federal Aviation Administration\\
MAPE &Mean Absolute Percentage Error\\
MAE &Mean Absolute Error\\
CUDA &Compute Unified Device Architecture\\
AMD &Advanced Micro Devices inc.\\
CPU &Central Processing Unit\\
RAM &Random Access Memory\\

### 2022-05-13 background.tex
Finished thesis. Implemented final feedback from instructor.

\chapter{Background\label{cha:background}}
In this chapter, we discuss the work our work builds on to give our work context. We cover the history of neural networks in machine learning, previous architectures that have been used to solve similar problems and the components of the solution we arrived at.

\section{Neural Networks}
Artificial Neural Networks (ANNs), generally known as just Neural Networks (NNs) were first theorized in the 1940s \cite{mcculloch_logical_1943}. They are supposed to emulate the function of biological brains. They are composed of \enquote{neurons} which, when they receive a signal, apply a function to the signal which they then relay to the next neuron. Neural networks are built in layers. A layer receives a tensor of input values, each neuron in the layer takes a weighted sum of the inputs and applies its activation function to the result. These values are then passed on to the next layer. Training a neural network is the process of optimizing the weights in the layers of the network such that a specified input yields the desired output via iteration to minimize the output of a loss function \cite{goodfellow_deep_2016}.\\
Frank Rosenblatt created the perceptron in the 1950s. A perceptron is a single layer of neurons with the ability to classify images a few hundred pixels in size. Rosenblatt's breakthrough idea was to train a neural network based on reinforcement from a dataset, strengthening successful connections and weakening unsuccessful connections. His training algorithm was very limited and so inefficient that it did not allow for neural networks with more than one layer \cite{rosenblatt_perceptron_1958}. Based on his work, Rosenblatt first theorized the ability of neural networks to recognize people and translate between languages \cite{vandeput_brief_nodate}.\\
In 1974 Paul Werbos proposed backpropagation to optimize neural networks \cite{Werbos:74}, which became popular in 1986  with the work of Rumelhart et al. \cite{rumelhart_learning_1986}.  Backpropagation is an efficient training algorithm opening up the possibility of training neural networks deeper than one layer using gradient descent. \enquote{The backpropagation algorithm works by computing the gradient of the loss function with respect to each weight by the chain rule, computing the gradient one layer at a time, iterating backward from the last layer to avoid redundant calculations of intermediate terms in the chain rule} \cite{noauthor_backpropagation_2022}. Backpropagation enabled the creation of deeper neural networks, notably used to read handwritten digits with good accuracy \cite{lecun_backpropagation_1989}.\\
Although backpropagation enabled neural networks with multiple layers, it did not enable significant depth. The most commonly used activation functions of the time, sigmoid and tanh, have derivatives close to zero, so stacking them flattens the gradient in the network values at an exponential rate. This slows down the gradient descent used to optimize the networks, also at an exponential rate. In the 2010s the ReLu activation function \cite{nair_rectified_2010} coupled with improved optimization algorithms like ADAM and massive advances in computer hardware opened the door for Deep Learning, neural networks with many more layers. Today, deep neural networks have been used to great effect in applications like Google DeepMind's AlphaGO \cite{silver_mastering_2016}, the first AI to beat the world champion in the game of go, and OpenAI's GPT3 model generates text human-like enough to pass the Turing test, to be confused with text written by a human \cite{vandeput_brief_nodate}.

\section{Recurrent Neural Networks\label{cha:RNN}}
Recurrent Neural Networks (RNNs), especially Gated Recurrent Units (GRUs) and Long Short-Term Memory (LSTM) cells \cite{Goodfellow-et-al-2016, hochreiter_long_1997} along with more advanced derivatives have been shown to be able to achieve good results when trained on previous weather and power generation data \cite{lin_temporal_2020, lee_forecasting_2018, jaidee_very_2019, su_machine_2019}.
RNNs are a class of deep neural networks where nodes are connected in a temporal sequence, allowing the network to learn time-based data efficiently \cite{lim_time_2021}. LSTMs and GRUs are types of RNNs in which, the nodes in the network have additional stored states controlled by the network via mechanisms that use time delays or feedback loops \cite{Goodfellow-et-al-2016, noauthor_recurrent_2021}. Figure~\ref{fig:RNNs} shows the structure of the nodes of the three different Recurrent Neural Network types discussed above.\\

\begin{figure}[ht!]
    \centering
    \subfloat[Plain Recurrent Neural Network (RNN)]{\includegraphics[scale=0.35]{imgs/RNN.png}\label{fig:RNN}}\qquad
    \subfloat[Gated Recurrent Unit (GRU)]{\includegraphics[scale=0.35]{imgs/GRU.png}\label{fig:GRU}}\qquad
    \subfloat[Long Short-Term Memory (LSTM)]{\includegraphics[scale=0.35]{imgs/LSTM.png}\label{fig:LSTM}}\qquad
    \caption{The cells of different Recurrent Neural Network types. $x_t$ represents inputs, $h_t$ and $C_t$ represent hidden layer vectors passed between cells. $\sigma$ and $tanh$ are activation functions. \cite{lopez_rnn_nodate}.
    \label{fig:RNNs}}
\end{figure}

@@

\afterpage{%
\begin{figure}
    \centering
    \includegraphics[scale=0.80]{imgs/transformer.png}
    \caption{The transformer architecture.\cite{vaswani_attention_2017}
    \label{fig:transformer}}
\end{figure}
\clearpage
}

@@

Transformers have been used for a while in Natural Language Processing (NLP) and have more recently been used very effectively in forecasting \cite{vaswani_attention_2017}. They can more effectively learn and preserve local context located far apart in the input data. As an example from NLP, the training data can contain multiple cases where time is being told, far apart. A transformer model would then more easily identify that the word \enquote{o’clock} is generally preceded by a number, whereas RNNs would be more likely to struggle with such distributed information. This can be applied to our work as similar weather patterns would reoccur days apart and cause a similar effect on measured irradiance at the power station \cite{lim_time_2021}. Figure~\ref{fig:transformer} shows an overview of the transformer architecture.

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{imgs/encoder_decoder.jpeg}
    \caption{The encoder/decoder structure \cite{nechu_what_2020}
    \label{fig:encoder_decoder}}
\end{figure}

\subsection{Positional encoding}
Unlike RNNs, the transformer architecture has no recurrence, or cell based memory, and does not inherently track the order of the input data, so positional information has to be injected into the network input for the model to be able to use the positional context of the data.\cite{vaswani_attention_2017}

@@

\section{Weather forecasts\label{cha:WRF}}
Numerical Weather Prediction (NWP) has used mathematical models of the oceans and atmosphere since the 1920s to predict the weather based on current conditions, although realistic results weren't achieved until the advent of computer simulations in the 1950s \cite{noauthor_numerical_2022}. The Weather Research and Forecasting (WRF) model is an NWP system created by the National Center for Atmospheric Research (NCAR), the National Oceanic and Atmospheric Administration (NOAA), the Air Force Weather Agency (AFWA), the Naval Research Laboratory (NRL), the University of Oklahoma (OU), and the Federal Aviation Administration (FAA) in the United States for both research and operational weather forecasting \cite{noauthor_weather_2022}.\\
WRF-SOLAR is an NWP model based on WRF specifically designed to model values useful for solar power, including high frequency irradiance calculations, more accurate solar position algorithms, and more robust aerosol and particle simulations with regard to radiation \cite{jimenez_wrf-solar_2016}.
\section{Previous work}
Lin et al. \cite{lin_temporal_2020} showed that while the more traditional GRUs and LSTMs show good performance on solar power prediction, Temporal Convolutional Neural Networks (TCNN) can show even better performance. \textit{\endquote{TCNN is a novel convolutional architecture designed for sequential modelling, which combines causal and dilated convolutions
and residual connections}} \cite{lin_temporal_2020}. \\
Jaidee et al. \cite{jaidee_very_2019} showed that LSTMs, GRUs and derivative methods show relatively similar performance for very short-term predictions of solar power production on the time scale of a few hours. This timescale is where it is critical to accurately predict cloud movements. 
These methods have been taken as far as building genetic algorithms to generate the optimal neural network for the purpose \cite{jaidee_very_2019}. These genetic algorithms operate on the same principles as animal breeding. They generate multiple candidate neural networks and use the best networks as a base to generate new networks for the next generation. This is done for multiple generations and in the end, the best neural network generated is used \cite{jaidee_very_2019}.\\
Su et al. \cite{su_machine_2019} did extensive experimentation on more conventional neural networks and newer, less known networks including  Non-linear Auto Regressive Neural Networks (NARXNN), as well as various statistical approaches. When predicting solar power output, NARXNN takes in the current power output values and considers them, in real time, as well as the past values of the power output of the system. These experiments showed that NARXNN had the best performance of these, by a rather large margin, and a hybrid method, of the better performing methods, showed even better results \cite{anderson_using_2018}. The statistical methods generally performed significantly worse in predicting the power output. \\
Satellite imagery has been used for predicting solar power output as well. A neural network was trained to predict solar power production, primarily from cloud cover information it learned from these satellite images \cite{jang_solar_2016}.\\

>>>>

\chapter{Background\label{cha:background}}
In this chapter, we discuss the work we build on. We cover the history of neural networks in machine learning, previous architectures that have been used to solve similar problems and the components of the solution we arrived at. We also give a summary of weather forecasting and its use in both our work and previous methods for solar power prediction.

\section{Neural Networks}
Artificial Neural Networks (ANNs), generally known as just Neural Networks (NNs) were first theorized in the 1940s \cite{mcculloch_logical_1943}. They are supposed to loosely emulate the function of biological brains. They are composed of \enquote{neurons} which, when they receive a signal, apply a function to the signal which they then relay to the next neuron. Neural networks are built in layers. A layer receives a tensor of input values, each neuron in the layer takes a weighted sum of the inputs and applies its activation function to the result. These values are then passed on to the next layer. Training a neural network is the process of optimizing the weights in the layers of the network such that a specified input yields the desired output via iteration to minimize the output of a loss function \cite{goodfellow_deep_2016}.

Frank Rosenblatt created the perceptron in the 1950s. A perceptron is a single layer of neurons with the ability to classify images a few hundred pixels in size. Rosenblatt's breakthrough idea was to train a neural network based on reinforcement from a dataset, strengthening successful connections and weakening unsuccessful connections. His training algorithm was very limited and so inefficient that it did not allow for neural networks with more than one layer \cite{rosenblatt_perceptron_1958}. Based on his work, Rosenblatt first theorized the ability of neural networks to recognize people and translate between languages \cite{vandeput_brief_nodate}.

In 1974 Paul Werbos proposed backpropagation to optimize neural networks \cite{Werbos:74}, which became popular in 1986  with the work of Rumelhart et al. \cite{rumelhart_learning_1986}.  Backpropagation is an efficient training algorithm opening up the possibility of training neural networks deeper than one layer using gradient descent. \enquote{The backpropagation algorithm works by computing the gradient of the loss function with respect to each weight by the chain rule, computing the gradient one layer at a time, iterating backward from the last layer to avoid redundant calculations of intermediate terms in the chain rule} \cite{noauthor_backpropagation_2022}. Backpropagation enabled the creation of deeper neural networks, notably used to read handwritten digits with good accuracy \cite{lecun_backpropagation_1989}.

Although backpropagation enabled neural networks with multiple layers, it did not enable significant depth. The most commonly used activation functions of the time, sigmoid and tanh, have derivatives close to zero, so stacking them flattens the gradient in the network values at an exponential rate. This slows down the gradient descent used to optimize the networks, also at an exponential rate. In the 2010s the ReLu activation function \cite{nair_rectified_2010} coupled with improved optimization algorithms like ADAM \cite{kingma_adam_2014} and massive advances in computer hardware opened the door for Deep Learning, neural networks with many more layers. Today, deep neural networks have been used to great effect in applications like Google DeepMind's AlphaGO \cite{silver_mastering_2016}, the first AI to beat the world champion in the game of Go, and OpenAI's GPT3 model generates text human-like enough to pass the Turing test, to be confused with text written by a human \cite{vandeput_brief_nodate}.

\section{Recurrent Neural Networks\label{cha:RNN}}
Recurrent Neural Networks (RNNs), especially Gated Recurrent Units (GRUs) \cite{cho_properties_2014} and Long Short-Term Memory (LSTM) cells \cite{hochreiter_long_1997, Goodfellow-et-al-2016} along with more advanced derivatives have been shown to be able to achieve good results when trained on previous weather and power generation data \cite{lin_temporal_2020, lee_forecasting_2018, jaidee_very_2019, su_machine_2019}.
RNNs are a class of deep neural networks where nodes are connected in a temporal sequence, allowing the network to learn time-based data efficiently \cite{lim_time_2021}. LSTMs and GRUs are types of RNNs in which, the nodes in the network have additional stored states controlled by the network via mechanisms that use time delays or feedback loops \cite{Goodfellow-et-al-2016, noauthor_recurrent_2021}. Figure~\ref{fig:RNNs} shows the structure of the nodes of the three different Recurrent Neural Network types discussed above.

\begin{figure}[ht!]
    \centering
    \subfloat[Plain Recurrent Neural Network (RNN)]{\includegraphics[scale=0.35]{imgs/RNN.png}\label{fig:RNN}}\qquad
    \subfloat[Gated Recurrent Unit (GRU)]{\includegraphics[scale=0.35]{imgs/GRU.png}\label{fig:GRU}}\qquad
    \subfloat[Long Short-Term Memory (LSTM)]{\includegraphics[scale=0.35]{imgs/LSTM.png}\label{fig:LSTM}}\qquad
    \caption{The cells of different Recurrent Neural Network types. $x_t$ represents inputs, $h_t$ and $C_t$ represent hidden layer vectors passed between cells. $\sigma$ and $tanh$ are activation functions. \cite{lopez_rnn_nodate}
    \label{fig:RNNs}}
\end{figure}

@@

\afterpage{%
\begin{figure}
    \centering
    \includegraphics[scale=0.80]{imgs/transformer.png}
    \caption{The transformer architecture. \cite{vaswani_attention_2017}
    \label{fig:transformer}}
\end{figure}
\clearpage
}

@@

Transformers have been used for some time in Natural Language Processing (NLP) and have more recently been used very effectively in forecasting \cite{vaswani_attention_2017}. They can more effectively learn and preserve local context located far apart in the input data. As an example from NLP, the training data can contain multiple cases where time is being told, far apart. A transformer model would then more easily identify that the word \enquote{o’clock} is generally preceded by a number, whereas RNNs would be more likely to struggle with such distributed information. This can be applied to our work as similar weather patterns would reoccur days apart and cause a similar effect on measured irradiance at the power station \cite{lim_time_2021}. Figure~\ref{fig:transformer} shows an overview of the transformer architecture.

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{imgs/encoder_decoder.jpeg}
    \caption{The encoder/decoder structure. \cite{nechu_what_2020}
    \label{fig:encoder_decoder}}
\end{figure}

\subsection{Positional encoding}
Unlike RNNs, the transformer architecture has no recurrence, or cell-based memory, and does not inherently track the order of the input data, so positional information has to be injected into the network input for the model to be able to use the positional context of the data.\cite{vaswani_attention_2017}

@@

\section{Weather forecasts\label{cha:WRF}}
Numerical Weather Prediction (NWP) has used mathematical models of the oceans and atmosphere since the 1920s to predict the weather based on current conditions, although realistic results weren't achieved until the advent of computer simulations in the 1950s \cite{noauthor_numerical_2022}. The Weather Research and Forecasting (WRF) model is an NWP system created by the National Center for Atmospheric Research (NCAR), the National Oceanic and Atmospheric Administration (NOAA), the Air Force Weather Agency (AFWA), the Naval Research Laboratory (NRL), the University of Oklahoma (OU), and the Federal Aviation Administration (FAA) in the United States for both research and operational weather forecasting \cite{noauthor_weather_2022}.

WRF-SOLAR is an NWP model based on WRF specifically designed to model values useful for solar power, including high-frequency irradiance calculations, more accurate solar position algorithms, and more robust aerosol and particle simulations with regard to radiation \cite{jimenez_wrf-solar_2016}.
\section{Previous work}
Lin et al. \cite{lin_temporal_2020} showed that while the more traditional GRUs and LSTMs show good performance on solar power prediction, Temporal Convolutional Neural Networks (TCNN) can achieve even better performance. \textit{\endquote{TCNN is a novel convolutional architecture designed for sequential modelling, which combines causal and dilated convolutions
and residual connections}} \cite{lin_temporal_2020}.

Jaidee et al. \cite{jaidee_very_2019} showed that LSTMs, GRUs and derivative methods show relatively similar performance for very short-term predictions of solar power production on the time scale of a few hours. This timescale is where it is critical to accurately predict cloud movements. 
These methods have been taken as far as building genetic algorithms to generate the optimal neural network for the purpose \cite{jaidee_very_2019}. These genetic algorithms operate on the same principles as animal breeding. They generate multiple candidate neural networks and use the best networks as a base to generate new networks for the next generation. This is done for multiple generations and in the end, the best neural network generated is used \cite{jaidee_very_2019}.

Su et al. \cite{su_machine_2019} did extensive experimentation on more conventional neural networks and newer, less known networks including  Non-linear Auto-Regressive Neural Networks (NARXNN), as well as various statistical approaches. When predicting solar power output, NARXNN takes in the current power output values and considers them, in real-time, as well as the past values of the power output of the system. These experiments showed that NARXNN had the best performance of these, by a rather large margin, and a hybrid method, of the better performing methods, showed even better results \cite{anderson_using_2018}. The statistical methods generally performed significantly worse in predicting the power output.

Satellite imagery has been used for predicting solar power output as well. A neural network was trained to predict solar power production, primarily from cloud cover information it learned from these satellite images \cite{jang_solar_2016}.

### 2022-05-13 discussion.tex
Small fixes
Finished thesis. Implemented final feedback from instructor.

\chapter{Discussion}
%In this section you discuss any issues that came up while developing
%the system.  If you found something particularly interesting,
%difficult, or an important learning experience, put it here.  This is
%also a good place to put additional figures and data.
Mean Absolute Percentage Error (MAPE) does not tell the whole story of a model's performance. The true measure of performance is how useful the model output is to energy producers. The model's confidence in a prediction is almost as important as the accuracy of the prediction, and the total energy produced over the day matters.
The forecast downward shortwave irradiance, the largest factor in measured irradiance, rarely drops. It is almost exclusively on very high variance days that it deviates from a sine wave. The downward diffuse shortwave irradiance, the other network input data value, is a much smaller part of the total irradiance and has, to the bare eye, seemingly no correlation with the observation values. Despite this, the models show generally fantastic performance in predicting the irradiance.
The full model was able to utilize the forecast data from the surrounding area to achieve noticeably better performance. Of course, since the limited model uses only one-ninth of the data of the full model, some hyperparameter tuning had to be done to optimize its performance and some of the performance difference could be explained by worse optimization.
With the data at hand, it is not possible to separate the performance of the weather forecast model producing our input data and the performance of our models. We do not know how close this model comes to maximizing the information available in the input data.

@@

\section{Limitations and Future Work}
We used a dataset of a fairly limited size, only 425 days. A few years of data for training the model would likely yield somewhat better results. The data can be processed more before being input into the network to account for known variables like daily and yearly swings in irradiance, which might improve performance. There exists satellite irradiance measurement data that might be used for improving accuracy. It would be interesting to see if a neural network could extract information from the seven pyranometers scattered across the power station, rather than only dropping the highest and lowest values and taking an average.

@@

Sadly, there is not a standard benchmark for solar irradiance forecast models, and the data used by others who have done similar work is not available, so we can't make a proper comparison between methodologies and models.

\section{Conclusion\label{sec:conclusions}}
The results clearly show that relying on forecast irradiance data from meteorological models is a very effective, but not perfect, method of accurately predicting irradiance in a single area. A transformer model, the Temporal Fusion Transformer, managed to very effectively translate patterns in meteorological irradiance forecasts to accurate local irradiance predictions utilizing spatially and temporally adjacent data.\\
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "DEGREE-NAME-YEAR"
%%% End:

>>>>

\chapter{Discussion}
%In this section you discuss any issues that came up while developing
%the system.  If you found something particularly interesting,
%difficult, or an important learning experience, put it here.  This is
%also a good place to put additional figures and data.
Mean Absolute Percentage Error (MAPE) does not tell the whole story of a model's performance. The true measure of performance is how useful the model output is to energy producers. The model's confidence in a prediction is almost as important as the accuracy of the prediction, and the total energy produced over the day matters.
The forecast downward shortwave irradiance, the largest factor in measured irradiance, rarely drops. It is almost exclusively on very high variance days that it deviates from a sine wave. The downward diffuse shortwave irradiance, the other network input data value, is a much smaller part of the total irradiance and has, to the bare eye, seemingly no correlation with the observation values. Despite this, the models show generally excellent performance in predicting the irradiance.
The full model was able to utilize the forecast data from the surrounding area to achieve noticeably better performance. Of course, since the limited model uses only one-ninth of the data of the full model, some hyperparameter tuning had to be done to optimize its performance and some of the performance difference could be explained by worse optimization.
With the data at hand, it is not possible to separate the performance of the weather forecast model producing our input data and the performance of our models. We do not know how close this model comes to maximizing the information available in the input data.

\pagebreak

@@

\section{Limitations and future work}
We used a dataset of a fairly limited size, only 425 days. A few years of data for training the model would likely yield somewhat better results. The data can be processed more before being input into the network to account for known variables like daily and yearly swings in irradiance, which might improve performance. There exists satellite irradiance measurement data that might be used for improving accuracy. It would be interesting to see if a neural network could extract information from the seven pyranometers scattered across the power station, rather than only dropping the highest and lowest values and taking an average.

@@

Sadly, there is not a standard benchmark for solar irradiance forecast models, and the data used by others who have done similar work is not available, so we can not make a proper comparison between methodologies and models, as weather predictability can vary hugely between locations. However, for some reference, Lee et al. produced a combination Convolutional Neural Network (CNN) and Long Short-Term Memory (CNN) model which managed to forecast the power output of a solar power station in Scotland 6 hours into the future with a MAPE of 25.85\% \cite{lee_forecasting_2018}.

\section{Conclusion\label{sec:conclusions}}
The results clearly show that relying on forecast irradiance data from meteorological models is an effective, but not perfect, method of accurately predicting irradiance in a single area. A transformer model, the Temporal Fusion Transformer, managed to effectively translate patterns in meteorological irradiance forecasts to accurate local irradiance predictions utilizing spatially and temporally adjacent data.\\
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "DEGREE-NAME-YEAR"
%%% End:
